---
title: "Macroeconomic and Financial Indicators: Cross-Country Exploratory Analysis"
author: "Haoran Li"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    theme: flatly
    fig_caption: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

```{r load-data, include=FALSE}
library(tidyverse)
library(ggcorrplot)
library(kableExtra)
library(patchwork)
library(scales)
library(WDI)

# ── Consistent visual theme ───────────────────────────────────────────────────
theme_report <- function() {
  theme_minimal(base_size = 11) +
    theme(
      strip.text      = element_text(face = "bold", size = 9),
      plot.title      = element_text(face = "bold", size = 11),
      plot.caption    = element_text(colour = "grey50", size = 8),
      legend.position = "bottom",
      legend.title    = element_blank()
    )
}

# ── Indicator codes and display labels ────────────────────────────────────────
indicators <- c(
  credit       = "FS.AST.PRVT.GD.ZS",
  gdp_growth   = "NY.GDP.MKTP.KD.ZG",
  gdp_pc       = "NY.GDP.PCAP.CD",
  investment   = "NE.GDI.TOTL.ZS",
  inflation    = "FP.CPI.TOTL.ZG",
  market_cap   = "CM.MKT.LCAP.GD.ZS",
  trade        = "NE.TRD.GNFS.ZS",
  unemployment = "SL.UEM.TOTL.ZS"
)

ind_cols <- names(indicators)   # short names used as column names throughout

var_display <- c(
  credit       = "Credit/GDP",
  gdp_growth   = "GDP Growth",
  gdp_pc       = "GDP per Capita",
  investment   = "Investment/GDP",
  inflation    = "Inflation",
  market_cap   = "Market Cap/GDP",
  trade        = "Trade/GDP",
  unemployment = "Unemployment"
)

# ── Fetch from World Bank API (cached to wdi_cache.rds after first run) ───────
# Delete wdi_cache.rds to force a refresh from the live API.
cache_file <- "wdi_cache.rds"

cache_valid <- file.exists(cache_file) &&
               all(ind_cols %in% names(readRDS(cache_file)))

if (cache_valid) {
  raw <- readRDS(cache_file)
} else {
  raw <- WDI(
    country   = "all",
    indicator = indicators,
    start     = 1960,
    end       = 2024,
    extra     = TRUE       # adds region, income, iso3c for filtering below
  )
  # Some WDI versions return full indicator codes instead of the short names
  # supplied in the indicators vector; rename those if present.
  codes_present <- indicators[unname(indicators) %in% names(raw)]
  if (length(codes_present) > 0) raw <- dplyr::rename(raw, !!!codes_present)
  saveRDS(raw, cache_file)
}

# ── Sovereign countries only (region == "Aggregates" marks composite rows) ────
ctry <- raw %>% filter(region != "Aggregates")

# ── WB income-group composites for time-series section ───────────────────────
income_groups <- c("High income", "Upper middle income",
                   "Lower middle income", "Low income")
income_wide <- raw %>% filter(country %in% income_groups)

# ── Rows with at least one non-NA indicator (used for sample-size reporting) ──
n_obs <- ctry %>% filter(if_any(all_of(ind_cols), ~ !is.na(.))) %>% nrow()

# ── Long formats: pivot indicator columns; recode variable to display labels ──
long_ctry <- ctry %>%
  select(country, iso3c, year, all_of(ind_cols)) %>%
  pivot_longer(all_of(ind_cols), names_to = "variable", values_to = "value") %>%
  mutate(variable = var_display[variable])

long_income <- income_wide %>%
  select(country, year, all_of(ind_cols)) %>%
  pivot_longer(all_of(ind_cols), names_to = "variable", values_to = "value") %>%
  mutate(variable = var_display[variable])
```

---

# Introduction

## Data source and collection

The dataset is drawn from the World Bank **World Development Indicators (WDI)** database, which aggregates national accounts, survey, and administrative data reported by member countries to the World Bank and other international organisations (IMF, ILO, OECD). Data are fetched programmatically via the `WDI` R package, which queries the World Bank API, ensuring the analysis is reproducible against the current database version.

## Unit of observation

Each observation represents a **country–year pair**. After excluding World Bank regional and income-group composites, the dataset contains `r format(n_obs, big.mark = ",")` country–year observations with at least one recorded value, spanning `r n_distinct(ctry$country)` economies over `r min(ctry$year)`–`r max(ctry$year)`.

## Variables, natural limits, and missing-value coding

All eight variables are continuous. Missing values are returned as `NA` by the WDI package. The table below lists each variable, its unit, and its natural limits—constraints that must hold for a value to be plausible.

```{r var-table}
var_meta <- tribble(
  ~Variable,         ~Unit,          ~`Natural Limits`,
  "Credit/GDP",       "% of GDP",    "≥ 0 (no strict upper bound)",
  "GDP Growth",       "%",           "No strict bounds; |x| > 50 is suspicious",
  "GDP per Capita",   "Current US$", "> 0",
  "Investment/GDP",   "% of GDP",    "≥ 0; typically < 60",
  "Inflation",        "%",           "No strict bounds; x > 500 suggests hyperinflation",
  "Market Cap/GDP",   "% of GDP",    "≥ 0 (no strict upper bound)",
  "Trade/GDP",        "% of GDP",    "≥ 0 (can exceed 100 for small open economies)",
  "Unemployment",     "%",           "0 ≤ x ≤ 100"
)

kable(var_meta,
      caption = "Table 1. Variable definitions, units, and natural limits.") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = FALSE, font_size = 12) %>%
  column_spec(3, italic = TRUE, width = "14em")
```

No variable is categorical. Natural limits that involve multiple variables simultaneously are not applicable here: each indicator is independently defined and there are no logical cross-variable constraints analogous to, say, a "non-smoker" coding cigarettes per day.

---

# Data Quality Checks

## Missing data

```{r missing-fig, fig.width = 8, fig.height = 3.5, fig.cap = "**Figure 1.** Left: percentage of country–year observations missing for each indicator. Right: share of individual countries with missing GDP Growth values, grouped by decade—illustrating that data coverage has improved markedly since the 1970s. Both panels are restricted to sovereign-country observations."}

# Panel A — missing rate by variable
miss_var <- long_ctry %>%
  group_by(variable) %>%
  summarise(pct = mean(is.na(value)) * 100, .groups = "drop") %>%
  mutate(variable = fct_reorder(variable, pct))

pA <- ggplot(miss_var, aes(pct, variable, fill = pct)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "#aec6cf", high = "#c0392b") +
  scale_x_continuous(labels = label_number(suffix = "%")) +
  labs(x = NULL, y = NULL, title = "A  Missing rate by variable") +
  theme_report()

# Panel B — temporal trend (GDP Growth as representative variable)
miss_time <- long_ctry %>%
  filter(variable == "GDP Growth") %>%
  mutate(decade = floor(year / 10) * 10) %>%
  group_by(decade) %>%
  summarise(pct = mean(is.na(value)) * 100, .groups = "drop")

pB <- ggplot(miss_time, aes(decade, pct)) +
  geom_line(colour = "#2980b9", linewidth = 1) +
  geom_point(colour = "#2980b9", size = 2) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(x = "Decade", y = NULL,
       title = "B  GDP Growth: missing rate by decade") +
  theme_report()

pA + pB
```

```{r missing-inline, include=FALSE}
worst_var <- miss_var %>% slice_max(pct, n = 1)
```

**`r worst_var$variable`** has the highest missing rate (`r round(worst_var$pct, 1)`%), because stock-market data exist only for economies with formal capital markets. Missingness is also strongly temporal: coverage before 1970 is sparse across all variables, reflecting limited statistical capacity in many newly independent states. Variables tend to co-absent: Market Cap/GDP, Credit/GDP, and Unemployment are commonly missing together, suggesting that financial-sector and labour-market reporting infrastructure develop jointly.

## Natural-limit violations and outliers

```{r limit-checks, include=FALSE}
unemp_viol <- ctry %>% filter(!is.na(unemployment),
                               unemployment < 0 | unemployment > 100)
gdppc_viol <- ctry %>% filter(!is.na(gdp_pc), gdp_pc <= 0)
hyperinfl  <- ctry %>%
  filter(!is.na(inflation), inflation > 500) %>%
  arrange(desc(inflation)) %>%
  select(country, year, inflation) %>%
  slice_head(n = 6)
```

No observation violates the unemployment bound \[0, 100\], and all GDP per Capita values are strictly positive—both limits are satisfied throughout. However, `r nrow(hyperinfl)` country–year observations record inflation above 500%; the most extreme cases correspond to well-documented hyperinflationary episodes.

```{r hyperinfl-table}
kable(
  hyperinfl %>% mutate(inflation = round(inflation, 1)),
  col.names = c("Country", "Year", "Inflation (%)"),
  caption   = "Table 2. Country–years with inflation exceeding 500%."
) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE,
                font_size = 12)
```

These values are arithmetically valid (e.g., Zimbabwe's 2007–2008 hyperinflation is well documented) and are retained in the dataset, but should be treated carefully—log-transformation or winsorisation is advisable before any regression analysis.

---

# Data Summaries

## Summary statistics

```{r summary-table}
fmt1 <- function(x) formatC(round(x, 1), format = "f", digits = 1, big.mark = ",")

summ <- ctry %>%
  select(all_of(ind_cols)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "v") %>%
  mutate(var = var_display[var]) %>%
  group_by(Variable = var) %>%
  summarise(
    N        = sum(!is.na(v)),
    Min      = fmt1(min(v,               na.rm = TRUE)),
    Q1       = fmt1(quantile(v, 0.25,    na.rm = TRUE)),
    Median   = fmt1(median(v,            na.rm = TRUE)),
    Mean     = fmt1(mean(v,              na.rm = TRUE)),
    Q3       = fmt1(quantile(v, 0.75,    na.rm = TRUE)),
    Max      = fmt1(max(v,               na.rm = TRUE)),
    `NA (%)` = round(mean(is.na(v)) * 100, 1),
    .groups  = "drop"
  ) %>%
  arrange(Variable)

kable(summ,
      caption = "Table 3. Summary statistics for all eight indicators (individual country–years).") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE, font_size = 11)
```

## Distributions

```{r dist-fig, fig.width = 8, fig.height = 4.5, fig.cap = "**Figure 2.** Distribution of each indicator across all country–year observations. GDP per Capita is displayed on a log₁₀ scale to handle its multi-order-of-magnitude range. Boxplot whiskers extend to 1.5 × IQR; individual points beyond are drawn in red."}

long_box <- ctry %>%
  select(all_of(ind_cols)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(
    display   = ifelse(var == "gdp_pc", log10(value), value),
    var_label = ifelse(var == "gdp_pc",
                       "GDP per Capita\n(log₁₀ US$)", var_display[var])
  )

ggplot(long_box, aes(x = factor(0), y = display)) +
  geom_boxplot(fill = "#3498db", alpha = 0.6,
               outlier.size = 0.4, outlier.colour = "#c0392b") +
  facet_wrap(~ var_label, scales = "free_y", ncol = 4) +
  labs(x = NULL, y = NULL) +
  theme_report() +
  theme(axis.text.x  = element_blank(),
        axis.ticks.x = element_blank())
```

GDP per Capita spans roughly three orders of magnitude (from under US\$200 to over US\$200,000), confirming the need for log-transformation in any cross-country regression. Inflation is strongly right-skewed with extreme outliers consistent with Table 2. The remaining six variables are broadly symmetric around their medians, though all carry pronounced right tails.

## Time-series trends by income group

```{r ts-fig, fig.width = 8, fig.height = 4, fig.cap = "**Figure 3.** Trends (1990–2024) for three indicators across World Bank income groups. Grey bands mark the Global Financial Crisis (2008–2009) and the COVID-19 shock (2020). Income-group series are World Bank composite aggregates and do not require individual-country imputation."}

focus_vars  <- c("GDP Growth", "Credit/GDP", "Trade/GDP")

recession_bands <- tibble(
  xmin = c(2007.5, 2019.5), xmax = c(2009.5, 2020.5),
  ymin = -Inf,            ymax = Inf
)

ts_data <- long_income %>%
  filter(variable %in% focus_vars,
         year >= 1990,
         !is.na(value)) %>%
  mutate(
    country  = factor(country,  levels = income_groups),
    variable = factor(variable, levels = focus_vars)
  )

ggplot(ts_data, aes(year, value, colour = country)) +
  geom_rect(data        = recession_bands,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            inherit.aes = FALSE, fill = "grey80", alpha = 0.5) +
  geom_line(linewidth = 0.75) +
  facet_wrap(~ variable, scales = "free_y", ncol = 3) +
  scale_colour_manual(values = c("#e74c3c", "#e67e22", "#27ae60", "#2980b9")) +
  scale_x_continuous(breaks = seq(1990, 2024, 10)) +
  labs(x = NULL, y = NULL) +
  theme_report()
```

Three patterns stand out. First, financial deepening (Credit/GDP) has risen persistently in high-income economies while remaining flat in low-income ones—a divergence that has widened since 2000. Second, trade openness (Trade/GDP) grew across all income groups through the 1990s–2000s, stalled after the 2008 crisis, and partially recovered, consistent with the slowdown in globalisation documented in the trade literature. Third, GDP growth rates synchronise sharply at both crisis episodes, with the COVID-19 contraction in 2020 being the deepest simultaneous shock across all income groups in the sample.

## Bivariate relationships

```{r corr-fig, fig.width = 5.5, fig.height = 5.2, fig.cap = "**Figure 4.** Spearman rank correlation matrix computed on pairwise-complete country–year observations. Spearman is used in preference to Pearson because GDP per Capita and Inflation are strongly right-skewed. Variables are reordered by hierarchical clustering. Crosses mark correlations not significant after Bonferroni correction (α = 0.05 / 28 pairs). Colour intensity reflects magnitude."}

# Compute on short names, then rename rows/cols for display before plotting
corr_mat <- ctry %>%
  select(all_of(ind_cols)) %>%
  cor(use = "pairwise.complete.obs", method = "spearman")

corr_p <- ctry %>%
  select(all_of(ind_cols)) %>%
  cor_pmat(method = "spearman")

# Inline values extracted before renaming
r_credit_gdppc  <- round(corr_mat["credit",     "gdp_pc"],   2)
r_mktcap_credit <- round(corr_mat["market_cap", "credit"],   2)
r_infl_growth   <- round(corr_mat["inflation",  "gdp_growth"], 2)

# Rename to display labels for the plot
rownames(corr_mat) <- colnames(corr_mat) <- var_display[colnames(corr_mat)]
rownames(corr_p)   <- colnames(corr_p)   <- var_display[colnames(corr_p)]

ggcorrplot(
  corr_mat,
  p.mat      = corr_p,
  sig.level  = 0.05 / 28,
  insig      = "pch",
  pch.cex    = 3,
  hc.order   = TRUE,
  type       = "lower",
  lab        = TRUE,
  lab_size   = 3,
  colors     = c("#c0392b", "white", "#2980b9")
) +
  theme_report() +
  theme(axis.text.x    = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        legend.title    = element_text(size = 9))  # restore: colour bar label is meaningful here
```

The strongest association is between Credit/GDP and GDP per Capita (ρ = `r r_credit_gdppc`), and between Market Cap/GDP and Credit/GDP (ρ = `r r_mktcap_credit`), confirming that financial depth and per-capita wealth cluster together—consistent with the financial-development literature. Inflation and GDP Growth are negatively correlated (ρ = `r r_infl_growth`), in line with evidence on the growth costs of macroeconomic instability. Trade/GDP is largely orthogonal to financial variables, reflecting that trade openness is driven more by geography and trade policy than by domestic financial development.

---

# Conclusions

This report examined eight World Bank macroeconomic indicators covering `r n_distinct(ctry$country)` economies from `r min(ctry$year)` to `r max(ctry$year)`. The principal data quality finding is that financial-sector variables—particularly Market Cap/GDP (`r round(filter(miss_var, variable == "Market Cap/GDP")$pct, 1)`% missing)—are systematically absent in lower-income economies, introducing a selection bias that researchers must address before fitting models that include these regressors. Distributional analysis reveals that GDP per Capita and Inflation are strongly right-skewed and span multiple orders of magnitude, making log-transformation a practical necessity rather than an optional refinement. Exploratory analysis highlights two structural patterns relevant to subsequent modelling: a persistent and widening divergence in financial depth across income groups since the 1990s, and a tight co-movement between financial variables and per-capita wealth, suggesting a latent financial development factor that could serve as a dimension-reduction target. Researchers extending this dataset towards causal inference should account for its two-way panel structure (country and year fixed effects), handle the non-random missing data in financial variables explicitly, and consider the documented hyperinflationary outliers as leverage points in any linear model.
