---
title: "Macroeconomic and Financial Indicators: Cross-Country Exploratory Analysis"
author: "Haoran Li"
output:
  pdf_document:
    toc: false
    number_sections: false
    latex_engine: xelatex
    fig_caption: true
geometry: "margin=2cm"
fontsize: 10pt
classoption: twocolumn
header-includes:
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{newunicodechar}
  - \newunicodechar{≥}{\ensuremath{\geq}}
  - \newunicodechar{≤}{\ensuremath{\leq}}
  - \newunicodechar{ρ}{\ensuremath{\rho}}
  - \newunicodechar{α}{\ensuremath{\alpha}}
  - \setlength{\dbltextfloatsep}{4pt plus 1pt minus 2pt}
  - \setlength{\textfloatsep}{4pt plus 1pt minus 2pt}
  - \setlength{\intextsep}{3pt plus 1pt minus 1pt}
  - \setlength{\floatsep}{4pt plus 1pt minus 1pt}
  - \setlength{\dblfloatsep}{4pt plus 1pt minus 1pt}
  - \setlength{\abovecaptionskip}{3pt}
  - \setlength{\belowcaptionskip}{1pt}
  - \renewcommand{\dbltopfraction}{0.9}
  - \renewcommand{\dblfloatpagefraction}{0.8}
  - \setcounter{dbltopnumber}{4}
  - \raggedbottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  out.width = "\\linewidth",  # scale all figures to text width in PDF
  dev       = "cairo_pdf"     # supports Unicode characters in plot text
)
```

```{r load-data, include=FALSE}
library(tidyverse)
library(ggcorrplot)
library(kableExtra)
library(patchwork)
library(scales)
library(WDI)

# ── Consistent visual theme ───────────────────────────────────────────────────
theme_report <- function() {
  theme_minimal(base_size = 9) +
    theme(
      strip.text      = element_text(face = "bold", size = 7.5),
      plot.title      = element_text(face = "bold", size = 9),
      plot.caption    = element_text(colour = "grey50", size = 7),
      legend.position = "bottom",
      legend.title    = element_blank(),
      legend.text     = element_text(size = 7.5),
      plot.margin     = margin(3, 4, 3, 4, "pt")
    )
}

# ── Indicator codes and display labels ────────────────────────────────────────
indicators <- c(
  credit       = "FS.AST.PRVT.GD.ZS",
  gdp_growth   = "NY.GDP.MKTP.KD.ZG",
  gdp_pc       = "NY.GDP.PCAP.CD",
  investment   = "NE.GDI.TOTL.ZS",
  inflation    = "FP.CPI.TOTL.ZG",
  market_cap   = "CM.MKT.LCAP.GD.ZS",
  trade        = "NE.TRD.GNFS.ZS",
  unemployment = "SL.UEM.TOTL.ZS"
)

ind_cols <- names(indicators)   # short names used as column names throughout

var_display <- c(
  credit       = "Credit/GDP",
  gdp_growth   = "GDP Growth",
  gdp_pc       = "GDP per Capita",
  investment   = "Investment/GDP",
  inflation    = "Inflation",
  market_cap   = "Market Cap/GDP",
  trade        = "Trade/GDP",
  unemployment = "Unemployment"
)

# ── Read pre-downloaded snapshot (Dataset.csv, retrieved 2026-02-24) ──────────
# The indicators were identified and verified via the WDI R package
# (see FindData.Rmd). This report uses the fixed CSV snapshot downloaded from
# World Bank DataBank on 2026-02-24, covering 1960-2025, to ensure results are
# fully reproducible without an internet connection.
series_to_col <- c(
  "Domestic credit to private sector (% of GDP)"                            = "credit",
  "GDP growth (annual %)"                                                    = "gdp_growth",
  "GDP per capita (current US$)"                                             = "gdp_pc",
  "Gross capital formation (% of GDP)"                                       = "investment",
  "Inflation, consumer prices (annual %)"                                    = "inflation",
  "Market capitalization of listed domestic companies (% of GDP)"            = "market_cap",
  "Trade (% of GDP)"                                                         = "trade",
  "Unemployment, total (% of total labor force) (modeled ILO estimate)"      = "unemployment"
)

raw <- read_csv("../Dataset.csv", na = c("..", "", "N/A"), show_col_types = FALSE) %>%
  rename(country = `Country Name`, code = `Country Code`, series = `Series Name`) %>%
  select(-`Series Code`) %>%
  mutate(col = series_to_col[series]) %>%
  filter(!is.na(col)) %>%
  pivot_longer(cols = matches("\\[YR"), names_to = "yr_raw", values_to = "value") %>%
  mutate(year = as.integer(str_extract(yr_raw, "\\d{4}"))) %>%
  select(country, code, col, year, value) %>%
  pivot_wider(names_from = col, values_from = value, values_fn = mean)

# ── Separate economies from World Bank aggregate groups ───────────────────────
# WDI_data$country ships with the WDI package; no internet call needed.
actual_iso3 <- WDI_data[["country"]] %>%
  filter(region != "Aggregates") %>%
  pull(iso3c)

ctry <- raw %>%
  filter(code %in% actual_iso3) %>%
  rename(iso3c = code)

# ── WB income-group composites for time-series section ───────────────────────
income_groups <- c("High income", "Upper middle income",
                   "Lower middle income", "Low income")
income_wide <- raw %>% filter(country %in% income_groups)

# ── Rows with at least one non-NA indicator (used for sample-size reporting) ──
n_grid <- nrow(ctry)   # full country-year grid including all-NA rows
n_obs  <- ctry %>% filter(if_any(all_of(ind_cols), ~ !is.na(.))) %>% nrow()

# ── Long formats: pivot indicator columns; recode variable to display labels ──
long_ctry <- ctry %>%
  select(country, iso3c, year, all_of(ind_cols)) %>%
  pivot_longer(all_of(ind_cols), names_to = "variable", values_to = "value") %>%
  mutate(variable = var_display[variable])

long_income <- income_wide %>%
  select(country, year, all_of(ind_cols)) %>%
  pivot_longer(all_of(ind_cols), names_to = "variable", values_to = "value") %>%
  mutate(variable = var_display[variable])
```

# Introduction

## Data source and collection

The data come from the World Bank **World Development Indicators (WDI)** database (World Bank, 2026), which compiles national accounts, survey, and administrative records submitted by member countries to the World Bank, IMF, ILO, and OECD. Indicator codes were checked against the `WDI` R package (full details in `FindData.Rmd`). Rather than pulling from the live API, this report uses a fixed snapshot (`Dataset.csv`) downloaded from [World Bank DataBank](https://databank.worldbank.org/source/world-development-indicators) on 24 February 2026—covering 1960–2025—so all results can be reproduced offline.

## Unit of observation

Each row is a **country–year pair**. After excluding World Bank regional and income-group composites, the full grid contains `r format(n_grid, big.mark = ",")` country–year slots across `r n_distinct(ctry$country)` economies over `r min(ctry$year)`–`r max(ctry$year)`; of these, `r format(n_obs, big.mark = ",")` have at least one non-missing indicator value. Missingness rates reported below use the full `r format(n_grid, big.mark = ",")` grid as the denominator, not the `r format(n_obs, big.mark = ",")` active-observation count. The sample covers UN member states and dependent territories that maintain separate national accounts (e.g., Hong Kong SAR, Virgin Islands U.S.), so "economy" is the more accurate label throughout.

## Variables, natural limits, and missing-value coding

All eight indicators are continuous numeric variables; the WDI package codes missing entries as `NA` rather than sentinel values like −99. Since no variable is categorical, checks for valid category membership and full category coverage are not applicable here. No cross-variable logical constraints apply either—each indicator is defined independently. Table 1 lists each variable alongside its unit and the natural limits any plausible value must satisfy.

```{r var-table}
var_meta <- tribble(
  ~Variable,        ~Description,                                              ~Unit,              ~`Natural Limits`,
  "Credit/GDP",      "Domestic credit to the private sector",                  "\\% of GDP",       "$\\geq 0$ (no strict upper bound)",
  "GDP Growth",      "Annual growth rate of real GDP",                         "\\%",              "No strict bounds; $|x| > 50$ is suspicious",
  "GDP per Capita",  "Gross domestic product divided by population",           "Current US\\$",    "$> 0$",
  "Investment/GDP",  "Gross capital formation (fixed investment + inventories)","\\% of GDP",      "No strict lower bound under SNA 2008; extreme negatives warrant scrutiny",
  "Inflation",       "Annual change in consumer price index",                  "\\%",              "No strict bounds; $x > 500$ suggests hyperinflation",
  "Market Cap/GDP",  "Market capitalisation of listed domestic companies",     "\\% of GDP",       "$\\geq 0$ (no strict upper bound)",
  "Trade/GDP",       "Sum of exports and imports of goods and services",       "\\% of GDP",       "$\\geq 0$ (can exceed 100 for entrepôt economies)",
  "Unemployment",    "Share of the labour force without work (ILO estimate)",  "\\%",              "$0 \\leq x \\leq 100$"
)

kbl(var_meta, booktabs = TRUE, longtable = FALSE,
    caption = "Variable definitions, units, and natural limits.",
    escape = FALSE,
    format = "latex") %>%
  kable_styling(latex_options = c("striped", "condensed", "HOLD_position"),
                full_width = FALSE, font_size = 7.5) %>%
  column_spec(1, width = "1.5cm") %>%
  column_spec(2, width = "2.3cm") %>%
  column_spec(3, width = "1.2cm") %>%
  column_spec(4, width = "2.6cm")
```

# Data Quality Checks

## Missing data

```{r missing-fig, fig.width = 7, fig.height = 2.8, fig.pos = "!t", fig.env = "figure*", fig.cap = "Left: percentage of country--year observations missing for each indicator. Right: share of economies with missing GDP Growth values, grouped by decade---illustrating that data coverage has improved markedly since the 1970s. Both panels are restricted to economy-level observations (aggregates excluded)."}

# Panel A — missing rate by variable
miss_var <- long_ctry %>%
  group_by(variable) %>%
  summarise(pct = mean(is.na(value)) * 100, .groups = "drop") %>%
  mutate(variable = fct_reorder(variable, pct))

pA <- ggplot(miss_var, aes(pct, variable, fill = pct)) +
  geom_col(show.legend = FALSE) +
  scale_fill_gradient(low = "#aec6cf", high = "#c0392b") +
  scale_x_continuous(labels = label_number(suffix = "%")) +
  labs(x = NULL, y = NULL, title = "A  Missing rate by variable") +
  theme_report()

# Panel B — temporal trend (GDP Growth as representative variable)
miss_time <- long_ctry %>%
  filter(variable == "GDP Growth") %>%
  mutate(decade = floor(year / 10) * 10) %>%
  group_by(decade) %>%
  summarise(pct = mean(is.na(value)) * 100, .groups = "drop")

pB <- ggplot(miss_time, aes(decade, pct)) +
  geom_line(colour = "#2980b9", linewidth = 1) +
  geom_point(colour = "#2980b9", size = 2) +
  scale_y_continuous(labels = label_number(suffix = "%")) +
  labs(x = "Decade", y = NULL,
       title = "B  GDP Growth: missing rate by decade") +
  theme_report()

pA + pB
```

```{r missing-inline, include=FALSE}
worst_var <- miss_var %>% slice_max(pct, n = 1)
```

```{r co-missing, include=FALSE}
co_miss_pct <- ctry %>%
  filter(is.na(market_cap)) %>%
  summarise(pct = mean(is.na(credit)) * 100) %>%
  pull(pct) %>% round(1)

co_miss_unemp <- ctry %>%
  filter(is.na(market_cap)) %>%
  summarise(pct = mean(is.na(unemployment)) * 100) %>%
  pull(pct) %>% round(1)

co_miss_invest <- ctry %>%
  filter(is.na(credit)) %>%
  summarise(pct = mean(is.na(investment)) * 100) %>%
  pull(pct) %>% round(1)
```

**`r worst_var$variable`** leads on missingness (`r round(worst_var$pct, 1)`%)—stock-exchange data do not exist for countries without formal capital markets—and pre-1970 coverage is thin across most variables. The gaps are correlated: `r co_miss_pct`% of observations missing Market Cap/GDP are also missing Credit/GDP; `r co_miss_invest`% missing Credit/GDP are also missing Investment/GDP. Financial variables form a coherent missing-data block and should be imputed or filtered as a group.

## Natural-limit violations and outliers

```{r limit-checks, include=FALSE}
# Unemployment: strict bounds [0, 100]
unemp_viol <- ctry %>% filter(!is.na(unemployment),
                               unemployment < 0 | unemployment > 100)

# GDP per Capita: must be strictly positive
gdppc_viol  <- ctry %>% filter(!is.na(gdp_pc), gdp_pc <= 0)
gdppc_min   <- ctry %>% filter(!is.na(gdp_pc)) %>%
  slice_min(gdp_pc, n = 1, with_ties = FALSE) %>% select(country, year, gdp_pc)
gdppc_max   <- ctry %>% filter(!is.na(gdp_pc)) %>%
  slice_max(gdp_pc, n = 1, with_ties = FALSE) %>% select(country, year, gdp_pc)

# Credit/GDP: flag extreme highs (>200%) as worth reviewing
credit_max  <- ctry %>% filter(!is.na(credit)) %>%
  slice_max(credit, n = 1, with_ties = FALSE) %>% select(country, year, credit)

# Investment/GDP: SNA 2008 allows negatives (inventory drawdowns), but flag them
invest_neg  <- ctry %>% filter(!is.na(investment), investment < 0)
invest_min  <- ctry %>% filter(!is.na(investment)) %>%
  slice_min(investment, n = 1, with_ties = FALSE) %>% select(country, year, investment)

# GDP Growth: flag |x| > 50 as suspicious
gdpg_extreme <- ctry %>% filter(!is.na(gdp_growth), abs(gdp_growth) > 50)
gdpg_max     <- ctry %>% slice_max(gdp_growth, n = 1, with_ties = FALSE) %>%
  select(country, year, gdp_growth)

# Trade/GDP: flag values > 400 (far beyond typical small open-economy range)
trade_extreme <- ctry %>% filter(!is.na(trade), trade > 400) %>%
  slice_max(trade, n = 1) %>% select(country, year, trade)

# Inflation: hyperinflation threshold 500%
hyperinfl <- ctry %>%
  filter(!is.na(inflation), inflation > 500) %>%
  arrange(desc(inflation)) %>%
  select(country, year, inflation) %>%
  slice_head(n = 6)

top_hyperinfl <- hyperinfl %>% slice_head(n = 1)

# Lower-tail zeros: Trade/GDP, Credit/GDP, Market Cap/GDP can be exactly 0 —
# flag count and one example for each to check whether they are genuine or coded missings
trade_zero      <- ctry %>% filter(!is.na(trade),      trade      == 0)
credit_zero     <- ctry %>% filter(!is.na(credit),     credit     == 0)
mktcap_zero     <- ctry %>% filter(!is.na(market_cap), market_cap == 0)
trade_zero_ex   <- trade_zero  %>% slice_head(n = 1)
credit_zero_ex  <- credit_zero %>% slice_head(n = 1)
```

**Unemployment and GDP per Capita** respect their natural limits throughout. The minimum GDP per Capita (US\$`r round(gdppc_min$gdp_pc, 1)`, `r gdppc_min$country` `r gdppc_min$year`) is an exchange-rate artefact; the maximum (US\$`r format(round(gdppc_max$gdp_pc), big.mark=",")`, `r gdppc_max$country` `r gdppc_max$year`) is plausible for a very small, very wealthy economy.

**Negative Investment/GDP**: `r nrow(invest_neg)` observations show negative gross capital formation (min `r round(invest_min$investment, 1)`%, `r invest_min$country` `r invest_min$year`). SNA 2008 permits this during large inventory drawdowns; values are retained but warrant scrutiny in investment-focused models.

**Extreme GDP Growth**: `r nrow(gdpg_extreme)` observations exceed $|x|>50\%$ (max `r round(gdpg_max$gdp_growth, 1)`%, `r gdpg_max$country` `r gdpg_max$year`). These fit post-conflict or oil-boom patterns—not recording errors—but carry high leverage in any regression.

**Extreme Trade/GDP**: Maximum `r round(trade_extreme$trade, 0)`% (`r trade_extreme$country`, `r trade_extreme$year`). Values above 400% are plausible for entrepôt economies where re-exports far exceed domestic output, making the GDP denominator misleadingly small.

**Extreme Credit/GDP**: Maximum `r round(credit_max$credit, 0)`% (`r credit_max$country`, `r credit_max$year`)—plausible for highly financialised economies (Hong Kong SAR, Cyprus) but a high-leverage point in credit regressions.

**Extreme Market Cap/GDP**: Ratios above 100% arise when multinationals list locally but operate abroad (e.g., mainland Chinese firms on the Hong Kong exchange); the inflated ratio is a denominator mismatch, not bad data.

**Lower-tail zeros**: The 0.0% minima shown for Trade/GDP, Credit/GDP, and Market Cap/GDP in Table 3 are display-rounding artefacts—the one-decimal-place format rounds small positive values down to 0.0. No observation is exactly zero for any of these variables (`r nrow(trade_zero)` exact-zero rows for Trade/GDP; `r nrow(credit_zero)` for Credit/GDP; `r nrow(mktcap_zero)` for Market Cap/GDP), so no miscoding correction is needed.

**Hyperinflation**: `r nrow(hyperinfl)` country–year observations record inflation above 500%.

```{r hyperinfl-table, results='asis'}
t2 <- kbl(
  hyperinfl %>% mutate(inflation = round(inflation, 1)),
  col.names = c("Country", "Year", "Inflation (%)"),
  caption   = "Country--years with inflation exceeding 500\\%.",
  booktabs  = TRUE, longtable = FALSE,
  format    = "latex"
) %>%
  kable_styling(latex_options = c("striped"),
                full_width = FALSE, font_size = 9)
# [!h]: place here in the text column, overriding topfraction limits;
# prevents the table from floating backward above the paragraph that introduces it
cat(sub("\\\\begin\\{table\\}", "\\\\begin{table}[t]", as.character(t2)))
```

Each entry in Table 2 is arithmetically valid and corresponds to a known currency crisis—the `r top_hyperinfl$country` episode in `r top_hyperinfl$year` being the most severe. These observations stay in the dataset, but log-transformation or winsorisation is strongly advisable before fitting any regression.

# Data Summaries

## Summary statistics

```{r summary-table}
fmt1 <- function(x) formatC(round(x, 1), format = "f", digits = 1, big.mark = ",")

summ <- ctry %>%
  select(all_of(ind_cols)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "v") %>%
  mutate(var = var_display[var]) %>%
  group_by(Variable = var) %>%
  summarise(
    N        = sum(!is.na(v)),
    Min      = fmt1(min(v,               na.rm = TRUE)),
    Q1       = fmt1(quantile(v, 0.25,    na.rm = TRUE)),
    Median   = fmt1(median(v,            na.rm = TRUE)),
    Mean     = fmt1(mean(v,              na.rm = TRUE)),
    Q3       = fmt1(quantile(v, 0.75,    na.rm = TRUE)),
    Max      = fmt1(max(v,               na.rm = TRUE)),
    `NA (%)` = round(mean(is.na(v)) * 100, 1),
    .groups  = "drop"
  ) %>%
  arrange(Variable)

kbl(summ, booktabs = TRUE, longtable = FALSE,
    caption = "Summary statistics for all eight indicators (individual country--years).",
    format = "latex") %>%
  kable_styling(latex_options = c("striped", "condensed", "scale_down", "HOLD_position"),
                full_width = FALSE, font_size = 8)
```

## Distributions

```{r dist-fig, fig.width = 3.3, fig.height = 5.5, fig.pos = "htbp", fig.env = "figure", fig.cap = "Distribution of each indicator across all country--year observations. GDP per Capita and Inflation are displayed on a log10 scale (positive values only) to handle extreme right skew; negative inflation observations are excluded from those panels only. Boxplot whiskers extend to 1.5 IQR; individual points beyond are drawn in red."}

long_box <- ctry %>%
  select(all_of(ind_cols)) %>%
  pivot_longer(everything(), names_to = "var", values_to = "value") %>%
  filter(!is.na(value)) %>%
  # Log-transform GDP per Capita (always positive) and Inflation (positive only)
  filter(!(var == "gdp_pc"   & value <= 0),
         !(var == "inflation" & value <= 0)) %>%
  mutate(
    display   = ifelse(var %in% c("gdp_pc", "inflation"), log10(value), value),
    var_label = case_when(
      var == "gdp_pc"    ~ "GDP per Capita\n(log10 US$)",
      var == "inflation" ~ "Inflation\n(log10 %)",
      TRUE               ~ var_display[var]
    )
  )

ggplot(long_box, aes(x = factor(0), y = display)) +
  geom_boxplot(fill = "#3498db", alpha = 0.6,
               outlier.size = 0.4, outlier.colour = "#c0392b") +
  facet_wrap(~ var_label, scales = "free_y", ncol = 2) +
  labs(x = NULL, y = NULL) +
  theme_report() +
  theme(axis.text.x  = element_blank(),
        axis.ticks.x = element_blank())
```

Log-transformation is necessary for two variables (Figure 2). GDP per Capita spans roughly three orders of magnitude, and Inflation's distribution still shows a pronounced right tail even on a log scale—a direct footprint of the hyperinflationary episodes in Table 2. The other six variables sit more symmetrically around their medians, though none are entirely free of outliers.

## Time-series trends by income group

```{r ts-fig, fig.width = 6.5, fig.height = 2.4, fig.pos = "!t", fig.env = "figure*", fig.cap = "Trends (1990--2023) for three indicators across World Bank income groups. Grey bands mark the Global Financial Crisis (2008--2009) and the COVID-19 shock (2020). Income-group series are World Bank composite aggregates and do not require individual-country imputation."}

focus_vars  <- c("GDP Growth", "Credit/GDP", "Trade/GDP")

recession_bands <- tibble(
  xmin = c(2007.5, 2019.5), xmax = c(2009.5, 2020.5),
  ymin = -Inf,            ymax = Inf
)

ts_data <- long_income %>%
  filter(variable %in% focus_vars,
         year >= 1990,
         !is.na(value)) %>%
  mutate(
    country  = factor(country,  levels = income_groups),
    variable = factor(variable, levels = focus_vars)
  )

ggplot(ts_data, aes(year, value, colour = country)) +
  geom_rect(data        = recession_bands,
            aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax),
            inherit.aes = FALSE, fill = "grey80", alpha = 0.5) +
  geom_line(linewidth = 0.75) +
  facet_wrap(~ variable, scales = "free_y", ncol = 3) +
  scale_colour_manual(values = c("#e74c3c", "#e67e22", "#27ae60", "#2980b9")) +
  scale_x_continuous(breaks = seq(1990, 2025, 10)) +
  labs(x = NULL, y = NULL) +
  theme_report()
```

Three patterns emerge (Figure 3). Credit/GDP has risen steadily in high-income economies since 2000 while stagnating in low-income ones—the gap keeps widening. Trade/GDP grew through the 1990s–2000s, stalled post-2008, and only partially recovered, consistent with Constantinescu, Mattoo & Ruta (2020). GDP growth synchronises sharply at both crisis windows; the 2020 contraction is the deepest in the sample.

## Bivariate relationships

```{r corr-fig, fig.width = 3.3, fig.height = 3.0, fig.pos = "htbp", fig.env = "figure", fig.cap = "Spearman rank correlation matrix computed on pairwise-complete country--year observations. Spearman is used in preference to Pearson because GDP per Capita and Inflation are strongly right-skewed. Variables are reordered by hierarchical clustering. Blank cells indicate correlations not reaching significance after Bonferroni correction ($\\alpha = 0.05 / 28$ pairs). Colour intensity reflects magnitude."}

# Compute on short names, then rename rows/cols for display before plotting
corr_mat <- ctry %>%
  select(all_of(ind_cols)) %>%
  cor(use = "pairwise.complete.obs", method = "spearman")

corr_p <- ctry %>%
  select(all_of(ind_cols)) %>%
  cor_pmat(method = "spearman")

# Inline values extracted before renaming
r_credit_gdppc  <- round(corr_mat["credit",     "gdp_pc"],     2)
r_mktcap_credit <- round(corr_mat["market_cap", "credit"],     2)
r_infl_growth   <- round(corr_mat["inflation",  "gdp_growth"], 2)

# Dynamic interpretation of the inflation–growth rank correlation
infl_growth_desc <- if (abs(r_infl_growth) < 0.05) {
  paste0(
    "show essentially no rank correlation (ρ = ", r_infl_growth, "), ",
    "suggesting that the classical negative inflation–growth link operates ",
    "primarily through extreme episodes rather than across the full distribution"
  )
} else if (r_infl_growth < 0) {
  paste0(
    "are negatively correlated (ρ = ", r_infl_growth, "), ",
    "consistent with evidence on the growth costs of macroeconomic instability"
  )
} else {
  paste0(
    "are weakly positively correlated (ρ = ", r_infl_growth, "), ",
    "a finding driven by high-growth emerging markets that experienced moderate ",
    "inflation during catch-up episodes; the classical negative relationship ",
    "documented in the literature emerges more clearly after conditioning on income level"
  )
}

# Rename to display labels for the plot
rownames(corr_mat) <- colnames(corr_mat) <- var_display[colnames(corr_mat)]
rownames(corr_p)   <- colnames(corr_p)   <- var_display[colnames(corr_p)]

ggcorrplot(
  corr_mat,
  p.mat      = corr_p,
  sig.level  = 0.05 / 28,
  insig      = "blank",
  hc.order   = TRUE,
  type       = "lower",
  lab        = TRUE,
  lab_size   = 1.8,
  colors     = c("#c0392b", "white", "#2980b9")
) +
  labs(x = NULL, y = NULL) +
  theme_report() +
  theme(axis.text.x    = element_text(angle = 45, hjust = 1, size = 6.5),
        axis.text.y    = element_text(size = 6.5),
        legend.position = "right",
        legend.title    = element_text(size = 7.5),
        legend.text     = element_text(size = 6.5),
        legend.key.size = unit(0.35, "cm"))
```

Figure 4 shows Credit/GDP and GDP per Capita as the tightest pair (ρ = `r r_credit_gdppc`), with Market Cap/GDP and Credit/GDP close behind (ρ = `r r_mktcap_credit`)—the familiar financial-depth cluster (King & Levine, 1993; Levine, 2005). These estimates are from a non-random ~19% subsample biased toward higher-income economies (N ~2,700 vs. N ~11,000 for non-financial pairs), so they may not generalise and some blank cells may reflect power limitations rather than true absence of association. Inflation and GDP Growth `r infl_growth_desc`. Trade/GDP is nearly uncorrelated with the financial cluster, consistent with trade openness being driven by geography and policy rather than domestic financial conditions.

# Conclusions

The dataset covers `r n_distinct(ctry$country)` economies from `r min(ctry$year)` to `r max(ctry$year)`. Its most consequential data quality issue is the non-random absence of financial variables—Market Cap/GDP is missing in `r round(filter(miss_var, variable == "Market Cap/GDP")$pct, 1)`% of country–years, concentrated in lower-income economies—a selection bias that must be made explicit before fitting any model that includes these regressors. Log-transformation is not optional for GDP per Capita or Inflation, both of which span multiple orders of magnitude; hyperinflationary observations (Table 2) additionally require winsorisation or case-by-case treatment. The most relevant structural pattern is the tight co-movement of financial depth with per-capita income (ρ > 0.7), forming a latent dimension recoverable by PCA or factor analysis, alongside a persistent widening of Credit/GDP across income groups since the 1990s. Researchers extending this work in the cross-country growth empirics tradition (Dollar & Kraay, 2002) should plan for two-way fixed effects, explicit missingness handling, and potential enrichment with institutional indicators (Acemoglu, Johnson & Robinson, 2001) to separate financial development from governance effects.

# References {-}

\footnotesize
\setlength{\parskip}{1pt}
\setlength{\parindent}{0pt}
\setlength{\hangindent}{1em}

Acemoglu, D., Johnson, S., & Robinson, J. A. (2001). The colonial origins of comparative development. *American Economic Review*, *91*(5), 1369--1401.

Constantinescu, C., Mattoo, A., & Ruta, M. (2020). The global trade slowdown: Cyclical or structural? *World Bank Economic Review*, *34*(1), 121--142.

Dollar, D., & Kraay, A. (2002). Growth is good for the poor. *Journal of Economic Growth*, *7*(3), 195--225.

King, R. G., & Levine, R. (1993). Finance and growth: Schumpeter might be right. *Quarterly Journal of Economics*, *108*(3), 717--737.

Levine, R. (2005). Finance and growth: Theory and evidence. In *Handbook of Economic Growth* (Vol. 1A, pp. 865--934). Elsevier.

World Bank. (2026). *World Development Indicators* [Dataset]. World Bank DataBank. Retrieved 24 February 2026.

\normalsize

